Seabird assemblages are linked to the major western boundary current off eastern Australia

### 2023-10-13 ms_preprint/ms_preprint.Rmd
{rticle} with the ms content updated

---
title: "Seabird assemblages are linked to the major western boundary current off eastern Australia"
author:
  - Nicholas W. Daudt:
      email: nicholaswdaudt@gmail.com
      institute: [MarSciUO]
      correspondence: true
  - Eric J. Woehler:
      institute: [ASG]
      correspondence: false
  - Matthew R. Schofield:
      institute: [MathStatUO]
      correspondence: false
  - Robert O. Smith:
      institute: [MarSciUO]
      correspondence: false
  - Leandro Bugoni:
      institute: [FURG]
      correspondence: false
  - William J. Rayment:
      institute: [MarSciUO]
      correspondence: false
institute:
  - MarSciUO: Department of Marine Science, University of Otago, Aotearoa New Zealand
  - ASG: Australasian Seabird Group, BirdLife Australia, Hobart, Australia
  - MathStatUO: Department of Mathematics and Statistics, University of Otago, Aotearoa New Zealand
  - FURG: Institute of Biological Sciences, Universidade Federal do Rio Grande--FURG, Rio Grande, RS, Brazil
abstract: |
  **Aim**: To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used 15 seabird at-sea ship-based survey data collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.
  **Results**: Based on 142,646 seabirds recorded, from 80 species, all models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at ~34°S, near the latitude at which the East Australian Current separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
  **Conclusion**: RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the rapid climatic changes that the East Australian Current is experiencing and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
output: 
  # rticles::sim_article:
  #   keep_tex: TRUE
  bookdown::word_document2:
    pandoc_args:
      - --lua-filter=../docs/scholarly-metadata.lua
      - --lua-filter=../docs/author-info-blocks.lua
      - --reference-doc=../docs/docx-template.docx 
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
---

<!-- YAML for bookdown+lua edited following: -->

<!-- https://www.r-bloggers.com/2019/08/rmarkdown-template-that-manages-academic-affiliations-docx-or-pdf-output/ -->

```{r setup, include = FALSE, echo = FALSE}
library(readr)
library(dplyr)
library(purrr)
library(gt)
```

\pagebreak

@@

In this study, we aim to identify assemblages of seabird species within the EAC and adjacent areas, and relate these assemblages to environmental conditions, by using at-sea surveys data and a state-of-the-art modelling technique. We use the Region of Common Profiles [RCP; @foster2013], a mixture model regarded as one of the best available tools for describing bioregions [@hill2020; @woolley2020]. RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics [@warton2015; @hill2020]. Eastern Australia supports a high diversity of seabirds and offers wide environmental gradients over a near 40° latitudinal range, from the tropics to the cool temperate (**ref?**) We hypothesised to find different seabird assemblages relating to different environmental conditions within the EAC. In addition, presence-absence and abundance data were used to test if the different numerical resolutions would result in the identification of similar assemblages. This study provides a repeatable framework on which future studies can compare to assess how future changes in the EAC may influence the distribution of seabird assemblages.

@@

```{r fig1-study-area, fig.cap = "**FIGURE 1.** Study area showing the main ocean currents (a), and seabird records made by season on top of the 1° latitude × 1° longitude grid cells (b). In (a), the East Australian Current (EAC) system is highlighted with its acronyms in bold and drawn in blue. The South Equatorial Current (SEC) and its branches, the North Caledonian Jet (NCJ) and South Caledonian Jet (SCJ), are drawn in red. The EAC 'eastern extension' is an eddy field (blue shaded area) often called the 'Tasman Front', from which the East Auckland Current originates (EAUC, in green). Figure (a) is adapted after Oke, Roughan, et al. (2019); for colour references see the online version.", fig.align = "center", echo = FALSE}

@@

data.frame(
  Variable = c("Sea Surface Temperature", "Sea Surface Temperature gradient", "Sea Surface Salinity", "Mixed Layer Depth", "8-day average Chlorophyll-a", "Log10 8-day average Chlorophyll-a", "Eddy Kinetic Energy", "Climatic mean EKE", "Climatic standard deviation EKE", "Climatic mean SSTgrad", "Bathymetry", "Slope", "Distance from coast"),
  Abbreviation = c("sst", "sst_grad", "sss", "mld", "chl8", "log10_chl8", "eke", "clim_eke_mean", "clim_eke_sd", "clim_sst_grad", "bat", "slope", "dist_coast"),
  Units = c("°Celsius", "delta-°Celsius", "PSU", "m", "mg/m^3", "log10(chl8)", "m/s", "mean(m/s)/season/10 years", "sd(m/s)/season/10 years", "mean(delta-°Celsius)/season/10 years", "m", "degrees", "km"),
  `Spatial resolution` = c("0.25° × 0.25°", "0.25° × 0.25°", "0.083° × 0.083°", "0.083° × 0.083°", "0.036° × 0.036°", "0.036° × 0.036°", "0.2° × 0.2°", "0.2° × 0.2°", "0.2° × 0.2°", "0.25° × 0.25°", "0.004° × 0.004°", "0.004° × 0.004°", "NA"),
  `Temporal resolution` = c("daily", "daily", "daily", "daily", "8-days composite", "8-days composite", "daily", "seasonal climatology", "seasonal climatology", "seasonal climatology", "NA", "NA", "NA"),
  Source = c("Huang et al. (2021)", "Derived from Huang et al. (2021)", "E.U. Copernicus Marine Service Information", "E.U. Copernicus Marine Service Information", "ERDDAP/NOAA", "Derived from ERDDAP/NOAA", "IMOS/AODN", "Derived from IMOS/AODN", "Derived from IMOS/AODN", "Derived from Huang et al. (2021)", "GEBCO", "Derived from GEBCO", "Calculated in R (see main text)")
) |>
  gt::gt() |>
  ## SST & SSTgrad
  gt::tab_footnote(footnote = 
                     gt::md("[NOAA OISST](https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html)"),
                   locations = gt::cells_body(columns = Source, rows = c(1:2, 10))) |>
  ## SSS & MLD
  gt::tab_footnote(footnote = 
                     gt::md("2011--2019: [GLOBAL_MULTIYEAR_PHY_001_030](https://doi.org/10.48670/moi-00021); 2020--2021: [GLOBAL_ANALYSIS_FORECAST_PHY_001_024](https://doi.org/10.48670/moi-00016)"),
                   locations = gt::cells_body(columns = Source, rows = 3:4)) |>
  ## CHL8 & log10-CHL8
  gt::tab_footnote(footnote = 
                     gt::md("[ERDDAP/NOAA](https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day.html)"),
                   locations = gt::cells_body(columns = Source, rows = 5:6)) |>
  ## EKE and derivations
  gt::tab_footnote(footnote = 
                     gt::md("[IMOS/AODN](https://portal.aodn.org.au/search), 'OceanCurrent - Gridded sea level anomaly - Delayed mode' data set"),
                   locations = gt::cells_body(columns = Source, rows = 7:9)) |>
  ## GEBCO & slope
  gt::tab_footnote(footnote = 
                     gt::md("[GEBCO](https://www.gebco.net/data_and_products/gridded_bathymetry_data/)"),
                   locations = gt::cells_body(columns = Source, rows = 11:12)) |>
  ## Improve cols names
  gt::cols_label(
    Variable = gt::md("**Variable**"),
    Abbreviation = gt::md("**Abbreviation**"),
    Units = gt::md("**Units**"),
    `Spatial.resolution` = gt::md("**Spatial resolution**"), 
    `Temporal.resolution` = gt::md("**Temporal resolution**"),
    Source = gt::md("**Source**")) |>
  ## Table title
  gt::tab_header(title = gt::md("**TABLE 1.** Environmental data used in the Region of Common Profile models applied to seabird data off eastern Australia, their spatial/temporal resolutions, units, and sources."))

```

@@

Data wrangling, modelling and visualization were done in R 4.2.0 [@r2022]; the full list of packages can be found in the Supplementary Material **XX**. Code and the modelling dataset (see \@ref(modellingdatasets)) are available in **Daudt et al. 2023 (Zenodo repo)**.

<!-- using packages 'plyr' 1.8.8 [@plyr], 'dplyr' 1.1.2 [@dplyr], 'tidyr' 1.3.0 [@tidyr], 'readr' 2.1.4 [@readr], 'tibble' 3.2.1 [@tibble], 'lubridate' 1.9.2 [@lubridate], 'stringr' 1.5.0 [@stringr], 'purrr' 1.0.1 [@purrr], 'ggplot2' 3.4.2 [@ggplot2], 'ggspatial' 1.1.7 [@ggspatial], 'patchwork' 1.1.2 [@patchwork], 'RColorBrewer' 1.1-3 [@rcolorbrewer], 'rnaturalearth' 0.3.2 [@rnaturalearth], 'sp' 1.6-0 [@sp1; @sp2], 'sf' 1.0-8 [@sf], 'mapview' 2.11.0 [@mapview], 'raster' 3.5-21 [@raster], 'terra' 1.6-7 [@terra], 'rerddap' 1.0.2 [@rerddap], 'rerddapXtracto' 1.1.4 [@rerddapXtracto], 'hadsstR' [@hadsstr], 'corrplot' 0.92 [@corrplot2021]. -->

@@

### Data {#modellingdatasets}

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling dataset for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (c. 110 $\times$ 110 km) and season (**Fig. 1b**), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were **standardised** before modelling.

@@

The model is specified in terms of assemblages and site-specific probabilities. There are $K$ assemblages (i.e. RCPs) that are common across the study area, where $K = (1, ..., k)$. Depending on whether the model is specified in terms of abundance or presence-absence, the assemblages are defined in terms of their structure at each $i$th site (a vector $y_{i}$), i.e. frequency or relative frequency of the species, respectively. The assemblage structure is denoted by $y_{i} = (y_{i1}, ... y_{iS})$, where $S$ is the number of species. The model estimates the probability of belonging to the $K$ assemblage given the site's assemblage structure. Thus, each site within the study area has a $k$-dimensional probability vector that gives the probability of the corresponding assemblage for that site. RCPs are a multivariate adaptation of a mixture-of-experts model; as such, the mixing proportions (assemblage structure) vary according to covariates (predictors) [@foster2013]. However, the final estimated assemblage structure is assumed to be constant for all sites belonging to the $K$ assemblage [@foster2013]. As such, RCPs can be directly interpreted as assemblages, and their species profiles as the assemblage structure.

@@

Due to seabird migration and dispersion, seasons could act as 'sampling artifacts' [@foster2017], thus biasing results [@ponti2023]. In light of the difference among seasons in the numbers of species that met our occurrence threshold (see \@ref(modellingdatasets)), we decided to use RCP models by season and data type (presence-absence and abundance). RCP models were used to relate the response variables (i.e. presence-absence or abundance of each seabird species in each grid cell) to the independent predictor covariates (i.e. averaged values of environmental variables in each cell). Presence-absence data were modelled based on a Bernoulli model, whereas abundance data based on a Negative Binomial model. RCPs were fitted using the 'ecomix' 1.0.0 package [@ecomix], with the following workflow: (i) find the best number of groups ($k$) that describes the data, (ii) re-fit the model by fixing $k$ and selecting optimal combination of covariates, (iii) check residuals, (iv) bootstrap the best model to estimate uncertainty around parameter estimates, and (v) produce results.

@@

We built a seasonal 11-year (2011--2021) average layer for predicting model results for each environmental covariate. These variables were then averaged within 1° latitude $\times$ 1° longitude grid cells, by season, between 9--47°S and 141--161°E. Note that 'clim_eke_mean' and 'clim_sst_grad' were processed beforehand as they may characterise persistent oceanographic features (described above) and thus represent the same value as 'eke' and 'sst_grad'. Data for prediction were **standardised** using the same mean and standard deviation from the modelling dataset.

@@

df <- 
  rbind(
    list.files(path = "../results/Bernoulli", pattern = "*04_full_and_best_models.csv", full.names = TRUE),
    list.files(path = "../results/NegBin", pattern = "*04_full_and_best_models.csv", full.names = TRUE)) %>% 
  purrr::map_df(~readr::read_csv(.)) %>% 
  dplyr::select(- `...1`) %>% 
  dplyr::mutate(data_type = factor(data_type, levels = c("bernoulli", "neg_bin")),
                season = factor(season, levels = c("summer", "autumn", "winter", "spring")),
                model = factor((ifelse(model == "full", "a", "b")), levels = c("a", "b")),
                bic = round(bic, digits = 2)) %>%
  dplyr::group_by(data_type) %>% 
  dplyr::arrange(data_type, season) %>% 
  as.data.frame(.) %>% 
  dplyr::ungroup(.)

# Correct a value in a cell
# Full abundance-autumn model did not fit properly, returning a value of -Inf
df[df$data_type=='neg_bin' & df$season=='autumn' & df$model=='a',]$bic <- "-Inf (n.f.)"

# A few tweaks to make the table look better
df$data_type <- c("Bernoulli", rep("", 7), "Negative Binomial", rep("", 7))
df$season <- rep(c("Summer", "", "Autumn", "", "Winter", "", "Spring", ""), 2)

gt::gt(df)|>
  ## Improve cols names
  gt::cols_label(
    data_type = gt::md("**Data type**"),
    season = gt::md("**Season**"),
    model = gt::md(" "),          ## To be excluded in the Word doc
    model_specification = gt::md("**Model specification**"), 
    bic = gt::md("**BIC**")) |>
  ## Table title
  gt::tab_header(title = gt::md("**TABLE 2.** Full (a) and best (b) Region of Common Profile Models applied to seabird data off eastern Australia, for both data types used (presence-absence, based on a Bernoulli model; abundance, based on a Negative Binomial model). Refer to Table 1 for the environmental data acronyms."))

rm("df")

## Can't figure out why this code below doesn't work...
# kableExtra::kbl(df, booktabs = T, align = "l", linesep = "") %>% 
#   kableExtra::column_spec(1, bold=T) %>%
#   kableExtra::collapse_rows(columns = c(1:3), 
#                             latex_hline = 'custom', 
#                             custom_latex_hline = 1:3, 
#                             row_group_label_position = 'stack')
```

```{r fig4-partial-plots, fig.cap = "**FIGURE 2.** Partial plots for the retained covariates in the spring (a) presence-absence and (b) abundance best models. The plot shows the fitted probability of belonging to a Region of Common Profiles (RCP) against the environmental value. All other predictors are held at their mean values. Results for other seasons and models can be found in the Supplementary Material (**Fig. S5**). Refer to Table 1 for the environmental data acronyms.", fig.align = "center", echo = FALSE}

@@

```{r fig2-point-pred, fig.cap = "**FIGURE 3.** Spatialized point predictions (hard-classes) for seabird assemblages (Region of Common Profile; RCP), by season, based on presence-absence (a) and abundance (b) models, off eastern Australia. Each grid was assigned to the RCP group with highest membership probability.", fig.align = "center", echo = FALSE}

knitr::include_graphics("../results/Fig2-point-predictions.png")
```

```{r fig3-prob-maps, fig.cap = "**FIGURE 4.** Predicted probability membership of each seabird assemblage (Region of Common Profiles; RCP) and grid, off eastern Australia, from spring presence-absence (a) and abundance (b) models. The central column, 'mean', corresponds to the point prediction and Bayesian boot-strapped, lower and upper confidence intervals (CI), on its sides. Results for other seasons and models can be found in the Supplementary Material (**Fig. S6**).", fig.align = "center", echo = FALSE}

knitr::include_graphics("../results/Fig3-probability-maps-spring.png")
```

@@

```{r fig5-spp-profiles, fig.cap = "**FIGURE 5.** Species profiles of seabird assemblages (Region of Common Profile; RCP) off eastern Australia in spring. Values are the average and 95% confidence intervals of probability of occurrence (a) and predicted mean abundance (b) for each species, based on 1000 Bayesian bootstraps. Values in (b) were log10-transformed to accommodate the high variation between species. Results for other seasons and models can be found in the Supplementary Material (**Fig. S7**).", fig.align = "center", echo = FALSE}

knitr::include_graphics("../results/Fig5-spp-profiles-spring.png")
```

@@

While remote tracking studies are advancing our understanding on how individual seabirds use their environment at fine scales [@tremblay2009], they fail to perceive how species assemblages interact with the environment and each other. Therefore, using at-sea counts of seabirds is a well-established method to capture these relationships [@ainley2012; @woehler2003; @hyrenbach2007]. Further, the spatial scales on which assemblages relate to their environment is variable and species-specific [@hunt1987; @ribic1997; @scales2017]. Understanding the biology of the studied group is thus essential to set the appropriate scales of analyses. Our choice of 1° latitude $\times$ 1° longitude was mainly determined by data availability reflecting survey effort, and the trade-off between the number of species per grid/season (see also \@ref(limitations)). Nonetheless, the chosen spatial scale is similar to other seabird distribution analyses conducted elsewhere [@harris2007; @santora2013; @sojitra2022]. Therefore, for describing seabird assemblages at a macro-scale, the chosen resolution seems to had been a good fit, accords with studies elsewhere, and the resultant models described patterns of seabird species' composition consistent with our current understanding of species' ecologies and life history strategies.

@@

## Limitations {#limitations}

@@

@@

>>>>

---
title: Seabird assemblages are linked to the major western boundary current off eastern Australia
authors:
  - name: Nicholas W. Daudt
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: nicholaswdaudt@gmail.com
  - name: Eric J. Woehler
    department: Australasian Seabird Group
    affiliation: BirdLife Australia
    location: Hobart, Australia
  - name: Matthew R. Schofield
    department: Department of Mathematics and Statistics
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
  - name: Robert O. Smith
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
  - name: Leandro Bugoni
    department: Institute of Biological Sciences
    affiliation: Universidade Federal do Rio Grande--FURG
    location: Rio Grande, RS, Brazil
  - name: William J. Rayment
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
abstract: |
  **Aim**: To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used 15 seabird at-sea ship-based survey data collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.
  **Results**: Based on 142,646 seabirds recorded, from 80 species, all models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at ~34°S, near the latitude at which the East Australian Current separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
  **Conclusion**: RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the rapid climatic changes that the East Australian Current is experiencing and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
keywords:
  - Assemblages
  - Community ecology
  - East Australian Current
  - Mixture models
  - Region of Common Profiles
bibliography: ../references.bib
# biblio-style: 
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
output: rticles::arxiv_article
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage{pdflscape}
- \DeclareUnicodeCharacter{0394}{\ensuremath{\Delta}}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
---

<!-- Running title: Seabird assemblages off East Australia -->
<!-- Note: I've changed the 'arxiv.sty' source file to include the running title in the PDF -->

```{r setup, include = FALSE, echo = FALSE}
library(readr)
library(dplyr)
library(purrr)
library(stringr)
library(kableExtra)

### If using {gt}:
# library(gt)
# options(gt.html_tag_check = FALSE)
```

@@

In this study, we aim to identify assemblages of seabird species within the EAC and adjacent areas, and relate these assemblages to environmental conditions, by using at-sea surveys data and a state-of-the-art modelling technique. We use the Region of Common Profiles [RCP; @foster2013], a mixture model regarded as one of the best available tools for describing bioregions [@hill2020; @woolley2020]. RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics [@warton2015; @hill2020]. Eastern Australia supports a high diversity of seabirds and offers wide environmental gradients over a near 40° latitudinal range, from the tropics to the cool temperate. We hypothesised to find different seabird assemblages relating to different environmental conditions within the EAC. In addition, presence-absence and abundance data were used to test if the different numerical resolutions would result in the identification of similar assemblages. This study provides a repeatable framework on which future studies can compare to assess how future changes in the EAC may influence the distribution of seabird assemblages.

@@

@@

@@

Data wrangling, modelling and visualization were done in R 4.2.0 [@r2022]; the full list of packages can be found in the Supplementary Material **XX**. Code and the modelling dataset (see \ref{sec:modellingdatasets}) are available in **Daudt et al. 2023 (Zenodo repo)**.

@@

### Data {#modellingdatasets}
\label{sec:modellingdatasets}

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling dataset for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (c. 110 $\times$ 110 km) and season (**Fig. 1b**), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

The model is specified in terms of assemblages and site-specific probabilities. There are $K$ assemblages (i.e. RCPs) that are common across the study area, $K = (1, ..., k)$. The outcome variable is a vector $y_{i} = (y_{i1}, ... y_{iS})$, that give the count or presence-absence of each of the $S$ species at site $i$. Depending on whether we model the counts or presence-absence data, the assemblages are defined in terms of the expected frequency or relative frequency of the species, respectively. An output of the model is the estimated probability that site $i$ has each of the $K$ possible assemblages. RCPs are a multivariate adaptation of a mixture-of-experts model; as such, the mixing proportions (assemblage structure) vary according to covariates (predictors) [@foster2013]. However, the final estimated assemblage structure is assumed to be constant for all sites belonging to the $K$ assemblage [@foster2013]. As such, RCPs can be directly interpreted as assemblages, and their species profiles as the assemblage structure.

@@

Due to seabird migration and dispersion, seasons could act as 'sampling artifacts' [@foster2017], thus biasing results [@ponti2023]. In light of the difference among seasons in the numbers of species that met our occurrence threshold (see \ref{sec:modellingdatasets}), we decided to use RCP models by season and data type (presence-absence and abundance). RCP models were used to relate the response variables (i.e. presence-absence or abundance of each seabird species in each grid cell) to the independent predictor covariates (i.e. averaged values of environmental variables in each cell). Presence-absence data were modelled based on a Bernoulli model, whereas abundance data based on a Negative Binomial model. RCPs were fitted using the 'ecomix' 1.0.0 package [@ecomix], with the following workflow: (i) find the best number of groups ($K$) that describes the data, (ii) re-fit the model by fixing $K$ and selecting optimal combination of covariates, (iii) check residuals, (iv) bootstrap the best model to estimate uncertainty around parameter estimates, and (v) produce results.

@@

We built a seasonal 11-year (2011--2021) average layer for predicting model results for each environmental covariate. These variables were then averaged within 1° latitude $\times$ 1° longitude grid cells, by season, between 9--47°S and 141--161°E. Note that 'clim_eke_mean' and 'clim_sst_grad' were processed beforehand as they may characterise persistent oceanographic features (described above) and thus represent the same value as 'eke' and 'sst_grad'. Data for prediction were standardised using the same mean and standard deviation from the modelling dataset.

@@

@@

@@

@@

While remote tracking studies are advancing our understanding on how individual seabirds use their environment at fine scales [@tremblay2009], they fail to perceive how species assemblages interact with the environment and each other. Therefore, using at-sea counts of seabirds is a well-established method to capture these relationships [@ainley2012; @woehler2003; @hyrenbach2007]. Further, the spatial scales on which assemblages relate to their environment is variable and species-specific [@hunt1987; @ribic1997; @scales2017]. Understanding the biology of the studied group is thus essential to set the appropriate scales of analyses. Our choice of 1° latitude $\times$ 1° longitude was mainly determined by data availability reflecting survey effort, and the trade-off between the number of species per grid/season (see also \ref{sec:limitations}). Nonetheless, the chosen spatial scale is similar to other seabird distribution analyses conducted elsewhere [@harris2007; @santora2013; @sojitra2022]. Therefore, for describing seabird assemblages at a macro-scale, the chosen resolution seems to had been a good fit, accords with studies elsewhere, and the resultant models described patterns of seabird species' composition consistent with our current understanding of species' ecologies and life history strategies.

@@

## Limitations
\label{sec:limitations}

@@

# Acknowledgements

We thank all observers who assisted in data collection, the crews of RV *Investigator* and RV *Falkor*, and the Australasian Seabird Group (BirdLife Australia) for logistical support. The authors would like to thank Dr. Skipton N.C. Woolley, `{ecomix}` developer, who kindly helped NWD build preliminary models and gave advice in several stages of model construction. NWD is supported by a Doctoral scholarship from the University of Otago.

@@

<div id="refs"></div>

\newpage

# Tables and Figures {-}

- **TABLE 1.** Environmental data used in the Region of Common Profile models applied to seabird data off eastern Australia, their spatial/temporal resolutions, units, and sources
- **TABLE 2.** Full (a) and best (b) Region of Common Profile Models applied to seabird data off eastern Australia, for both data types used (presence-absence, based on a Bernoulli model; abundance, based on a Negative Binomial model). Refer to Table 1 for the environmental data acronyms
- **FIGURE 1.** Study area showing the main ocean currents (a), and seabird records made by season on top of the 1° latitude × 1° longitude grid cells (b). In (a), the East Australian Current (EAC) system is highlighted with its acronyms in bold and drawn in blue. The South Equatorial Current (SEC) and its branches, the North Caledonian Jet (NCJ) and South Caledonian Jet (SCJ), are drawn in red. The EAC 'eastern extension' is an eddy field (blue shaded area) often called the 'Tasman Front', from which the East Auckland Current originates (EAUC, in green). Figure (a) is adapted after Oke, Roughan, et al. (2019); for colour references see the online version
- **FIGURE 2.** Partial plots for the retained covariates in the spring (a) presence-absence and (b) abundance best models. The plot shows the fitted probability of belonging to a Region of Common Profiles (RCP) against the environmental value. All other predictors are held at their mean values. Results for other seasons and models can be found in the Supplementary Material (Fig. S5). Refer to Table 1 for the environmental data acronyms
- **FIGURE 3.** Spatialized point predictions (hard-classes) for seabird assemblages (Region of Common Profile; RCP), by season, based on presence-absence (a) and abundance (b) models, off eastern Australia. Each grid was assigned to the RCP group with highest membership probability
- **FIGURE 4.** Predicted probability membership of each seabird assemblage (Region of Common Profiles; RCP) and grid, off eastern Australia, from spring presence-absence (a) and abundance (b) models. The central column, 'mean', corresponds to the point prediction and Bayesian boot-strapped, lower and upper confidence intervals (CI), on its sides. Note that, in case of two RCPs, RCP2 probablility equals to (1 - RCP1 probability). Results for other seasons and models can be found in the Supplementary Material (Fig. S6)
- **FIGURE 5.** Species profiles of seabird assemblages (Region of Common Profile; RCP) off eastern Australia in spring. Values are the average and 95% confidence intervals of probability of occurrence (a) and predicted mean abundance (b) for each species, based on 1000 Bayesian bootstraps. Values in (b) were log10-transformed to accommodate the high variation between species. Results for other seasons and models can be found in the Supplementary Material (Fig. S7)

\blandscape

data.frame(
  Variable = c("Sea Surface Temperature", "Sea Surface Temperature gradient", "Sea Surface Salinity", "Mixed Layer Depth", "8-day average Chlorophyll-a", "Log10 8-day average Chlorophyll-a", "Eddy Kinetic Energy", "Climatic mean EKE", "Climatic standard deviation EKE", "Climatic mean SSTgrad", "Bathymetry", "Slope", "Distance from coast"),
  # Abbreviation needs the `str_replace_all` because I'm using `escape=F` in `kbl()`
  Abbreviation = stringr::str_replace_all(c("sst", "sst_grad", "sss", "mld", "chl8", "log10_chl8", "eke", "clim_eke_mean", "clim_eke_sd", "clim_sst_grad", "bat", "slope", "dist_coast"), "_", "\\\\_"), 
  Units = c("°Celsius", paste0(parse(text = "\u0394"), "°Celsius"), "PSU", "m", "mg/m${^3}$", "$log_{10}$(chl8)", "m/s", "mean(m/s)/season/10 years", "sd(m/s)/season/10 years", paste0("mean(", paste0(parse(text = "\u0394"), "°Celsius"), ")/season/10 years"), "m", "degrees (°)", "km"),
  `Spatial resolution` = c("0.25° × 0.25°", "0.25° × 0.25°", "0.083° × 0.083°", "0.083° × 0.083°", "0.036° × 0.036°", "0.036° × 0.036°", "0.2° × 0.2°", "0.2° × 0.2°", "0.2° × 0.2°", "0.25° × 0.25°", "0.004° × 0.004°", "0.004° × 0.004°", "NA"),
  `Temporal resolution` = c("daily", "daily", "daily", "daily", "8-days composite", "8-days composite", "daily", "seasonal climatology", "seasonal climatology", "seasonal climatology", "NA", "NA", "NA"),
  Source = c("Huang et al. (2021)${^a}$", "Derived from Huang et al. (2021)${^a}$", "E.U. Copernicus Marine Service Information${^b}{^c}$", "E.U. Copernicus Marine Service Information${^b}{^c}$", "ERDDAP/NOAA${^d}$", "Derived from ERDDAP/NOAA${^d}$", "IMOS/AODN${^e}$", "Derived from IMOS/AODN${^e}$", "Derived from IMOS/AODN${^e}$", "Derived from Huang et al. (2021)${^a}$", "GEBCO${^f}$", "Derived from GEBCO${^f}$", "Calculated in R (see main text)")
) |>
  kableExtra::kbl(caption = " ", booktabs = TRUE, escape = FALSE)  %>% 
  kableExtra::row_spec(0, bold = TRUE) %>% 
  kableExtra::column_spec(1, width = "4cm") %>% 
  kableExtra::column_spec(6, width = "3cm") %>% 
  kableExtra::footnote(
    alphabet = c(
    "NOAA OISST: (https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html); ", 
    "2011--2019: GLOBAL_MULTIYEAR_PHY_001_030 (https://doi.org/10.48670/moi-00021); ", 
    "2020--2021: GLOBAL_ANALYSIS_FORECAST_PHY_001_024 (https://doi.org/10.48670/moi-00016); ",
    "ERDDAP/NOAA (https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day.html); ",
    "IMOS/AODN (https://portal.aodn.org.au/search), 'OceanCurrent - Gridded sea level anomaly - Delayed mode' data set; ",
    "GEBCO (https://www.gebco.net/data_and_products/gridded_bathymetry_data/)"))
